On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-23 tex/draft/draft4.tex
edits to absract: reorganize and mention analytic model

\title{On the Minimum Core Mass for Giant Planet Formation}

@@

The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, such as the location in the disk, nebular gas composition, or the accretion rate of solids. In this study we consider the limiting case of atmosphere growth around a core of fixed mass, in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the envelope. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within typical disk life timescales for a variety of disk conditions. We are primarily interested in the outer regions of the protoplanetary disk, where our assumptions of spherical symmetry and dust grains opacity are valid, and where the increasing number of directly imaged giant planets helps constrain formation models. The absolute minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for an ideal diatomic gas with a solar composition and a standard ISM dust opacity. Lower opacity and disk temperature significantly reduce the critical core mass, while a decrease in the mean molecular weight of the nebular gas results in a larger critical core mass. Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. As additional heating due to planetesimal accretion leads to slower cooling, our results represent an absolute minimum for the core mass needed to form a giant planet before disk dissipation.

%Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs. This core mass is typically quoted to be around $10 M_{\oplus}$. In static calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.

>>>>

\title{On the Minimum Core Mass for Giant Planet Formation at Wide Separations}

@@

The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. This value, however, depends on several factors, such as the location in the disk, atmospheric opacity, and the accretion rate of solids.  Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
Our model investigates the growth of atmospheres around solid cores of fixed mass.  Without planetesimal accretion to heat the atmosphere, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction. We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk.  For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for a solar composition atmosphere with a standard ISM dust opacity.   This trend is mainly explained by the cooler temperatures in the outer disk. Lower pressure and densities in the outer disk have a negligible effect on core mass requirements.  At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the atmosphere reduces the critical core mass.   The ongoing accretion of solids would delay or prevent atmospheric runaway growth.  Thus our low values for minimum core masses do not contradict the existence of ice giants, but emphasize the importance of the accretion history of solids.  We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.  Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim 20\%$ as massive as the core.
